R Notebook

### 2021-02-11 examples/Toy-example-balanced-accessibility.Rmd
Reviewed new disparity analysis and made small edits
Added an example of reaggregation of the accessibility indicator and an example of a bivariate choropleth map

The disparity for SoBi is shown in the following map. The figure reveals the spatial distribution in disparity, with levels of access that are lower than the mean in the periphery of the service area, compared to levels of access that are greater than the mean in the higher-density and more connected urban center. It is worth noting that levels of access that are greater than the mean are predominantly concentrated around the University in west Hamilton. Areas that have less than their equitable share of the level of service are some of the most socioeconomically disadvantaged areas in the city. Our analysis builds upon a previous and recent study [@hosfordWhoArePublic2018] which found that disadvantaged areas in Hamilton are better served by the SoBi system, compared to other Canadian cities [i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau] where advantaged areas have greater access. The authors [@hosfordWhoArePublic2018] note "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas. Thus, relative to other Canadian programs, Hamilton’s bicycle share program could more easily be established in lower income areas while still being near high population density areas". This is an artifact of the 2001 amalgamation of Hamilton with four surrounding communities.	The urban core, which is Hamilton proper, remains predominantly disadvantaged compared to the 4 amalgamated and less densely populated areas (e.g., Dundas, Ancaster, Stoney Creek, Waterdown) that surround the city. Interestingly, Dundas is an advantaged area but also has lower accessibility that is less than their equitable share. However, by calculating the disparity index, we have shown that more advantaged areas within the SoBi service area have higher levels of service and exceed its equitable share of service. 
# Disparity of system with ERI hubs

@@

# Disparity of original system without ERI hubs

@@

>>>>

The disparity for SoBi is shown in the following map. The figure reveals the spatial distribution in disparity, with levels of access that are lower than the mean in the periphery of the service area, compared to levels of access that are greater than the mean in the higher-density and more connected urban center. It is worth noting that levels of access that are greater than the mean are predominantly concentrated around the University in west Hamilton. Areas that have less than their equitable share of the level of service are some of the most socioeconomically disadvantaged areas in the city. Our analysis builds upon a previous and recent study [@hosfordWhoArePublic2018] which found that disadvantaged areas in Hamilton are better served by the SoBi system, compared to other Canadian cities [i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau] where advantaged areas have greater access. The authors [@hosfordWhoArePublic2018] note "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas. Thus, relative to other Canadian programs, Hamilton’s bicycle share program could more easily be established in lower income areas while still being near high population density areas". This is an artifact of the 2001 amalgamation of Hamilton with four surrounding communities.	The urban core, which is Hamilton proper, remains predominantly disadvantaged compared to the 4 amalgamated and less densely populated areas (e.g., Dundas, Ancaster, Stoney Creek, Waterdown) that surround the city. Interestingly, Dundas is an advantaged area but also has lower accessibility that is less than their equitable share. However, by calculating the disparity index, we have shown that more advantaged areas within the SoBi service area have higher levels of service and exceed its equitable share of service.

## Disparity of system with ERI hubs

@@

## Disparity of original system without ERI hubs

@@

## Reaggregating accessibility

One advantage of the Balanced FCA approach is that accessibility can be reaggregated while preserving the population and level of service. Here, we reaggregate the population to DAs. This allows us to then join the DA-level accessibility to DA-level census data for further analysis.

Read the DA-level data:
```{r}
load("hamilton_da_2016.RData")
```

Plot DAs:
```{r}
ggplot(data = hamilton_da_2016) +
  geom_sf()
```

Project the Hamilton DA zones to the same projection of the SoBi Service Area:
```{r}
hamilton_da_2016 <- st_transform(hamilton_da_2016, crs = st_crs(sobi_service))
```

Bounding box of DAs in serviced areas:
```{r}
bounding_box <- st_bbox(sobi_service %>%
                          st_buffer(500))
```

Extract DAs in service area:
```{r}
hamilton_da_sobi <- hamilton_da_2016[st_as_sfc(bounding_box),]
```

Plot accessibility and extracted DAs:
```{r}
ggplot() +
  geom_sf(data = population_sobi_original,
          aes(color = accessibility_o)) +
  geom_sf(data = hamilton_da_sobi,
          fill = NA) + 
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1)
```

Spatial join:
```{r}
accessibility_da_original <- population_sobi_original %>%
  st_join(hamilton_da_sobi %>% dplyr::select(DA))
```

Aggregate population and accessibility by DA:
```{r}
accessibility_da_original <- accessibility_da_original %>%
  group_by(DA) %>%
  summarize(population = sum(population),
            accessibility_o = sum(accessibility_o),
            .groups = "drop")

Check that accessibility (LOS) is preserved:
```{r}
sum(accessibility_da_original$accessibility_o)
sum(population_sobi_original$accessibility_o)
```

Check that the population is preserved:
```{r}
sum(accessibility_da_original$population)
sum(population_sobi_original$population)
```

Ok!

Add the geometry of DAs to the table `accessibility_da` and convert into simple features:
```{r}
accessibility_da_original <- accessibility_da_original %>%
  st_drop_geometry() %>%
  left_join(hamilton_da_sobi %>% 
              dplyr::select(DA, geometry),
            by = "DA") %>%
  st_as_sf()
```

Plot DA-level accessibility:
```{r}
ggplot() +
  geom_sf(data = accessibility_da_original,
          aes(fill = accessibility_o),
          color = "black") +
  scale_fill_distiller(palette = "YlOrRd", 
                        direction = 1)
```

Same thing for the other accessibility estimates. Spatial join:
```{r}
accessibility_da <- population_sobi %>%
  st_join(hamilton_da_sobi %>% dplyr::select(DA))
```

Aggregate population and accessibility by DA:
```{r}
accessibility_da <- accessibility_da %>%
  group_by(DA) %>%
  summarize(population = sum(population),
            accessibility = sum(accessibility),
            .groups = "drop")

Check that accessibility (LOS) is preserved:
```{r}
sum(accessibility_da$accessibility)
sum(population_sobi$accessibility)
```

Check that the population is preserved:
```{r}
sum(accessibility_da$population)
sum(population_sobi$population)
```

Ok!

Add the geometry of DAs to the table `accessibility_da` and convert into simple features:
```{r}
accessibility_da <- accessibility_da %>%
  st_drop_geometry() %>%
  left_join(hamilton_da_sobi %>% 
              dplyr::select(DA, geometry),
            by = "DA") %>%
  st_as_sf()
```

Plot DA-level accessibility:
```{r}
ggplot() +
  geom_sf(data = accessibility_da,
          aes(fill = accessibility),
          color = "black") +
  scale_fill_distiller(palette = "YlOrRd", 
                        direction = 1)
```

Join two accessibility estimates:
```{r}
accessibility_da <- accessibility_da %>%
  left_join(accessibility_da_original %>% 
              st_drop_geometry() %>%
              rename(population_o = population),
              dplyr::select(DA, accessibility_o, population_o),
            by = "DA") %>%
  mutate(accessibility_o = replace_na(accessibility_o, 0), # The population serviced grows with the new stations, and as a result some original population is zero
         population_o = replace_na(population_o, 0)) # Same for accessibility

Calculate the difference in accessibility with and without ERI hubs:
```{r}
accessibility_da <- accessibility_da %>%
  mutate(R_a = accessibility - accessibility_o)
```

Plot accessibility differences between with and without ERI hubs:
```{r}
ggplot() +
  geom_sf(data = accessibility_da,
          aes(fill = R_a),
          color = "black") +
  scale_fill_distiller(palette = "YlOrRd", 
                        direction = 1)
```

Retrieve census data using cancensus package (requires API key!)
```{r}
library(cancensus)
```

**NOTE TO ELISE**

You can get an API to use `cancensus` from [CensusMapper](https://censusmapper.ca/). Once you have an API key, you can store it in your system this way:
```{r eval=FALSE}
set_api_key("your-key", install = TRUE)
```

Once you have your API you can use `cancensus` like in this example.

Use `cancensus` to check and download census data. Check the regions:
```{r}
list_census_regions('CA16') %>% 
  filter(level == "CMA", name %in% c("Hamilton"))
```

We need PR_UID 35.

List census data sets:
```{r}
list_census_datasets()
```

We need CA16.

```{r}
find_census_vectors("median household income", 
                    dataset = "CA16", 
                    type = "total", 
                    query_type = "keyword", 
                    interactive = FALSE)
```

We need v_CA16_2397.

Retrieve data:
```{r eval=FALSE}
mhi <- get_census(dataset='CA16', 
                          regions=list(CMA=c("35537")),
                          vectors=c("v_CA16_2397"),
                          level='DA', 
                          use_cache = FALSE)
```

Join median household income to accessibility by DA:
```{r}
accessibility_da <- accessibility_da %>%
  left_join(mhi %>%
              transmute(DA = GeoUID,
                        median_household_income = `v_CA16_2397: Median total income of households in 2015 ($)`),
            by = "DA")
```

Summary of median household income:
```{r}
summary(accessibility_da$median_household_income)
```

There is a missing value (likely a censored unit).

Bivariate choropleth maps of accessibility and income:
```{r}
library(biscale)
library(cowplot)
```

Create a class for bivariate plotting (drop DA with missing income data):
```{r}
data <- bi_class(accessibility_da %>%
                   drop_na(),
                 x = accessibility_o, 
                 y = median_household_income, 
                 style = "quantile", 
                 dim = 3) %>%
  st_as_sf()
```

Plot:
```{r}
 map <- ggplot() +
  geom_sf(data = data, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "DkBlue", dim = 3) +
  bi_theme()

map
```

Create legend:
```{r}
legend <- bi_legend(pal = "DkBlue",
                    dim = 3,
                    xlab = "Higher accessibility ",
                    ylab = "Higher income",
                    size = 6)
```

Finalize plot:
```{r}
 cowplot::ggdraw() +
  cowplot::draw_plot(map, 0, 0, 1, 1) +
  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
```
